---
title: "HYD-RESPONSES Code Examples"
author: "Christoph von Matt"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This documents provides an exploratory overview of the main data products provided
within the HYD-RESPONSES dataset (see source). The dataset consists of daily catchment-level
time series data extracted from gridded hydro-meteorological datasets provided by
MeteoSwiss, SLF and ECMWF. Further derived indicators related to snow, water balance,
streamflow and deficit accumulation (CWD, SPI, SMRI, SPEI) are provided.

The code examples provide a guidance on:

* how to read-in the data
* how to combine data products
* how to select specific variables
* an example use and visualization of data

## Loading required libraries
```{r libraries, echo=TRUE}
# use package manager
if(!require("pacman")) install.packages("pacman")

# load required libraries
pacman::p_load("ggplot2", "readr", "dplyr", "forcats", "tidyr", "lubridate",
                    "purrr", "stringr", "tibble", "sf")

```

## Folder structure

Data is stored in the folder `data` and contains subfolders for each product and
shapefiles for both catchment outlets and outlines.

Available subfolders are:

* `01_timeseries`: Time series data
* `02_climatology`: Climatological information for the reference period 1991-2020
* `03_descriptive`: Information on catchment characteristics (climatological, terrestrial, ...)

The ESRI-file format only allows for a limited number of characters for column names.
Both catchment outlets and outlines thus only contain the geometries and catchment-identifiers (`ctchmnt`) and no additional information to minimize information loss in column names. Renaming to `catchmentnr` is suggested for cross-data compatibility.

```{r folder_structure}
# data directory
dir_data <- "../data/"
list.files(dir_data)

# read-in catchment outlets
catchment_outlets <- read_sf(file.path(dir_data, "catchment_outlets.shp"))
catchment_outlets

# rename for compatibility
catchment_outlets <- catchment_outlets %>% 
  rename(catchmentnr = ctchmnt)

```

## Read-in data products

Starting from catchment outlets (or catchment outlines), data products can be loaded and combined via the catchment identifiers.

All data files are stored as deliminator-separated csv-files (;) as river and place
name character strings contain commas.

```{r data_products}
# data directories
dir_timeseries <- file.path(dir_data, "01_timeseries")
dir_climatology <- file.path(dir_data, "02_climatology")
dir_descriptors <- file.path(dir_data, "03_descriptive")
```

### Catchment characteristics

It is generally a good idea to complement the catchment identifiers with the general station information which contain basic catchment characteristics (e.g., river and place name,
catchment area, mean height, glaciation percentage) but also information on the **station-specific** streamflow time series availability. Data availability is provided
for the full (`start_date`) , continuous (`start_date_continuous`) and homogeneous (`start_date_homogeneous`). For catchments where neither breakpoints nor gaps were detected,
the availability is identical for all.

```{r descriptors}

# check available files
list.files(dir_descriptors)

# complement catchment identifiers
catchment_outlets <- catchment_outlets %>% 
  left_join(
    read_delim(
      file.path(dir_descriptors, "HYDRESPONSES_descriptive_general_station_information.csv")
    )
  )

# information on data availability
catchment_outlets %>% 
  dplyr::select(catchmentnr, water_name, place, contains(c("start_date", "end_date"))) %>% 
  st_drop_geometry()

```

### Time series data

Time series are available for the following products:

* `01_base_variables`: basic (non-altered) hydro-meteorological variables extracted from gridded products
* `02_derived_variables`: information derived from basic variables on snow ($\Delta$SWE), water balance (P-E) and streamflow (7d average streamflow) and combinations thereof
* `03_anomalies`: standardized anomalies (z-scores: (value - mean)/sd)) for all variables across products and for multiple scales (daily, monthly: 31d, seasonal: 91d)
* `04_standardized_indices`: Time series of SPI, SPEI and SMRI
* `05_cumulative_deficits`: cumulative deficits of (potential) water balance and streamflow time series
* `06_events_nrs`: Time series of numbered streamflow drought events based on variable and fixed threshold definitions (values below threshold)



```{r timeseries_overview}

# check available files
list.files(dir_timeseries)

```

Monthly standardized anomalies are loaded for example catchment *2034 - Broye, Payerne (Caserne d' aviation)* which will be used for demonstration in the remaining part of this guide on the HYD-RESPONSES dataset. Note that only the data is only used for the homogeneous streamflow period.

```{r timeseries}

# check available files
list.files(file.path(dir_timeseries, "03_anomalies", "zscores")) %>%
  head(10)

# select example catchment
catchment_2034 <- catchment_outlets %>% 
  filter(catchmentnr == "2034")

# load the monthly (31d = scale_31) anomalies and filter only for the homogeneous streamflow period
data_anomalies <- read_delim(
  file.path(
    dir_timeseries, "03_anomalies", "zscores", "HYDRESPONSES_zscores_scale_31_catchment_2034.csv"
  )
) %>% 
  filter(date >= catchment_2034$start_date_homogeneous)

```

### Climatology

Each catchment is classified by a specific streamflow regime type which is available in the
already loaded general catchment characteristics (`FOEN_regime_type`). For catchment 2034, the characteristic streamflow regime is characterized as **pluvial inférieur**. Streamflow regime types are classified by considering the annual streamflow climatology.

Data on climatologies for all time series products are provided both DOY-based (for each day of the year, folder "DOY_movingwindow") and regular. DOY-based climatologies were assessed by using a moving window centered on the specific day for the specific scale (e.g., monthly = day - 15 to day + 15). Regular climatologies are provided per month, season, extended season and yearly.

To get an impression of how the characteristic streamflow regime type looks like, we first load the regular climatology and visualize the monthly mean streamflow levels.

(This may take some seconds longer as climatologies provide a comprehensive set of statistics...)

```{r climatology}

# check available files
list.files(file.path(dir_climatology, "regular")) %>% 
  head(10)

# load regular climatology
data_clim <- read_delim(
  file.path(
    dir_climatology, "regular", "HYDRESPONSES_climatology_regular_catchment_2034.csv"
  )
)
data_clim
```


The files containing climatological information contain climatologies for all scales and the
desired scale has to be selected before use (`time_scale`). Note that depending on scale, either the columns month, season or extended_season contain the information on the specific category. For a monthly climatology this corresponds to the month (1--12), for season to *DJF* (winter), *MAM* (spring), *JJA* (summer) and *SON* (autumn) and for extended_season *summer* (May -- October) and *winter* (November -- March).

Here, we are only interested in the monthly streamflow (*Q*) climatology.


```{r climatology_monthly}

# filter monthly climatology
data_clim_monthly <- data_clim %>% 
  filter(time_scale == "monthly")


# visualize the monthly streamflow climatology
ggplot(data_clim_monthly) +
  # visualize the IQR
  geom_ribbon(
    aes(month, ymin = Q_q25, ymax = Q_q75),
    alpha = 0.5, color = "grey60"
  ) +
  # visualize the monthly median streamflow
  geom_line(
    aes(month, Q_med)
  ) +
  scale_x_continuous("Month", breaks = 1:12) +
  scale_y_continuous("m3/s") +
  labs(
    # use catchment descriptors for title
    title = str_c(
      catchment_2034$catchmentnr, " - ", catchment_2034$water_name, " / ",
      catchment_2034$place, 
      " (", catchment_2034$FOEN_regime_type, ")"
    )
  ) +
  coord_fixed(ratio = 0.5)

```

### Streamflow drought events

Time series on events defined by fixed and variable threshold definitions are provided for streamflow droughts (`hydrological_events`) and for cumulative deficits for both streamflow (*CQWD*) and (potential) water balance (*CWD* and *CPWD*). For the subsequent analysis we are interested in streamflow droughts phases.

```{r eventseries}
# check streamflow drought files
list.files(file.path(dir_timeseries, "06_events_nrs", "hydrological_events")) %>% 
  head(10)

# load streamflow drought event time series
data_droughts <- read_delim(
  file.path(
    dir_timeseries, "06_events_nrs", "hydrological_events", "HYDRESPONSES_eventseries_hydroevents_catchment_2034.csv"
  )
)

```

Streamflow drought event series are available for variable and fixed threshold definitions.
Variable thresholds include monthly (`perc_31`) and seasonal (`perc_91`) percentiles. The only fixed threshold is the `Q347` which corresponds roughly to a yearly 5th-percentile derived based on the flow duration curve (FDC).

Here we use the yearly Q347-based streamflow drought series derived from 7 daily average streamflow (M7Q) series (`M7Q_Q347_0`).

Each event definition has two columns. In columns with `_event_nr`, drought phases are successively numbered while in columns with `_event_days_cumulative` days below the threshold are successively numbered per drought phase.

Example:    
NA NA NA 1 1 1 NA NA 2 2 NA NA (`event_nr`)  
NA NA NA 1 2 3 NA NA 1 2 NA NA (`event_days_cumulative`)


```{r eventseries_hydro}
# available event series
data_droughts %>% 
  colnames()

# select events based on a yearly fixed threshold definition (Q347)
data_droughts <- data_droughts %>% 
  dplyr::select(catchmentnr, date, contains("M7Q_Q347")) %>% 
  # again: filter homogeneous time series
  filter(date >= catchment_2034$start_date_homogeneous)

```


## Example streamflow drought analyis

In the last section, potential use cases and combination possibilities are demonstrated by analysing the temperature, precipitation and evaporation anomalies during streamflow drought events for the catchment *2034 - Broye, Payerne*.

### Combining datasets

The previously loaded anomalies can easily be combined with the streamflow drought event series as both all time series products share the columns `catchmentnr` and `date`.

Note that for some datasets additional variables may be required for combination which in some cases must be manually created. For example to add DOY-based climatologies, both datasets must also contain the variable `doy` (day of year) which can be added via `mutate(doy = lubridate::yday(date))`).

```{r combination}

# combine anomalies with event series
data_combined <- data_anomalies %>% 
  left_join(data_droughts)

```

### Composite analysis: Hydro-meteorological indicators during streamflow droughts

The time series now contain information on both the "normality" (or anomalousness) of the hydro-meteorological conditions as well as the dates where streamflow droughts occured. An interesting question now be if hydro-meteorological conditions differ between drought and non-drought phases. Further, hydro-meteorological conditions may also be more extreme for longer streamflow droughts.

Here, we compare values between drought and non-drought phases for monthly (31d) temperature and evaporation anomalies. Note that temperature and evaporation originate from different gridded datasets. Here the data availability is longest for the streamflow time series (`start_date_homogeneous = 1924-12-31`). The HYD-RESPONSES data provides ERA5-LAND data from 1950 onwards. The MeteoSwiss TabsD product (daily average temperature) has the shortest data availability and starts in 1961. Events can therefore only be analysed after 1961.

Boxplots for both temperature and evaporation show higher values during streamflow droughts and also boxplot-notches do not overlap!

```{r events_nonevents}

data_combined %>% 
  # ensure data coverage for all products
  filter(date >= dmy("01-01-1961")) %>% 
  dplyr::select(
    # keep information on drought phases
    contains(c("_event_nr", "event_days_cumulative")),
    # select monthly anomalies of temperature and evaporation
    TabsD_mean_standardized_anomaly_31,
    e_mean_mean_standardized_anomaly_31
  ) %>% 
  rename(
    event_nr = M7Q_Q347_0_event_nr,
    event_days_cumulative = M7Q_Q347_0_event_days_cumulative,
    Temperature = TabsD_mean_standardized_anomaly_31,
    Evaporation = e_mean_mean_standardized_anomaly_31
  ) %>% 
  # create a helper variable for drought / non-drought
  mutate(
    drought = !is.na(event_nr)
  ) %>% 
  # pivot for plotting
  pivot_longer(
    c("Temperature", "Evaporation")
  ) %>% 
  # visualize anomaly values
  ggplot() +
  geom_boxplot(
    aes(name, value, color = drought),
    notch = TRUE
  ) +
  ylab("Anomaly") + 
  xlab("Variable") +
  scale_color_manual(
    "Drought", 
    values = c("darkgreen", "tan2"), 
    labels = c("no", "yes")
  ) +
  labs(title = "Anomalies during streamflow drought") +
  coord_fixed(ratio = 0.2)
  

```

You have reached the end of this tutorial! 

More detailed information can be found in the HYD-RESPONSES documentation (Link). 

Thank you for reading and have fun with the HYD-RESPONSES dataset! :-)
